Novel genetic control of migratory diapause in Australian Monarch Butterflies

### 2023-04-24 paper/paper_draft.Rmd
edits for submission

---
title: "Novel genetic control of migratory diapause in Australian Monarch Butterflies"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  word_document:
    reference_docx: "paper-files/style_ref.docx"
  pdf_document:
    keep_tex: true
    fig_caption: true
    latex_engine: pdflatex
    fig_width: 7
    fig_height: 6
bibliography: "paper-files/citations.bib"
csl: "paper-files/council-of-science-editors-author-date.csl"
header-includes:
  - \usepackage[left]{lineno}
  - \linenumbers
  - \usepackage{setspace}\doublespacing
  - \usepackage{sectsty}\sectionfont{\fontsize{12}{12}\selectfont}
  - \usepackage{sectsty}\subsectionfont{\normalfont\itshape\fontsize{12}{12}\selectfont}
  - \usepackage[round]{natbib}
indent: true
sansfont: Times New Roman
fontsize: 12pt
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(snpR); library(data.table); library(ggplot2); library(gridExtra); library(dplyr); library(openxlsx); library(imager); library(nlme)

@@

# Methods
## Background
Although the ancestral range of monarch butterflies is believed to be in North and Central America, they have fairly recently expanded into South America and the Caribbean, and very recently to several locations throughout the Atlantic and Pacific [@oplerFieldGuideWestern1999; @Zhan2014]. In North America, the species is well known for its unique, multi-generational migratory life history wherein individuals dispersing northward in the spring do so over three to four generations and then return to their wintering grounds in a single step [@malcolmSpringRecolonizationEastern1993; @cockrellTimeTemperatureLatitudinal1993].

@@

## Sample Collection, Incubation, and Phenotyping
```{r, read_sample_meta_get_numbers, echo=FALSE,message=FALSE,include=FALSE,warning=FALSE}
sample.meta <- read.table("../data/sample_metadata.txt", header = T, stringsAsFactors = F)
old.data <- data.table::fread("../data/Aus_2018_metadata.txt", header = T)
old.data[,fam := as.numeric(gsub("_.+", "", ID))]

@@

In order to determine if any genomic regions were significantly associated with reproductive diapause, we used the snpR R package to filter out loci significantly out of Hardy-Weinburg Equilibrium ($p \le 0.000001$) in either year in order to help remove any paralogous loci and interpolated missing data by conducting binomial draws at missing genotypes using the minor allele frequencies at those loci [@hemstromSnpRUserFriendly2022]. To assess the basic level of genetic diversity amongst our samples, we then calculated the average number of pairwise differences ($\pi$), observed heterozygosity ($H_{O}$), and $F_{IS}$ according to Weir and Cockerham [-@Weir1984] for all of our samples and for each year independently using snpR [@hemstromSnpRUserFriendly2022]. We also calculated $F_{ST}$ according to Weir and Cockerham [-@Weir1984] between sample years and between both sample year and reproductive phenotype jointly [@hemstromSnpRUserFriendly2022]. We bootstrapped a significance level for $F_{ST}$ by bootstrapping individuals among populations with replacement 100 times. To determine if there was any substantial unexpected genetic structuring (other than family groups) in our samples, we conducted a Principal Component Analysis using snpR [@hemstromSnpRUserFriendly2022]. Since some individuals escaped their rearing dishes (but not the incubator) during rearing, we also confirmed the parentage of our samples using the Colony2 program, assuming polygamous, random mating [@Jones2010].

@@

# plot
Figure3b <- plot_manhattan(dat, "gmmat_pval_likely_pheno", chr = "scaffold", significant = 0.00005, suggestive = 0.0005, log.p = T,
                           chr_order = c(paste0("chr", 1:30), "other"), highlight_style = "red") # highlight top snp
Figure3b <- Figure3b$plot + ylab("-log10(p-value)") + xlab("Position") + theme(strip.text = element_blank())
```
In order to determine the degree to which diapause status was biased between families, we constructed a Fisher's exact test for independence using 2000 simulations in R version 4.2.2 [@rcoreteamLanguageEnvironmentStatistical2022]. We then conducted a Genome-Wide Association Study (GWAS) in order to determine if any loci were significantly associated with reproductive status. Since our samples were composed of many groups of full and half-siblings and came from two genetically different years, we used the GMMAT R package to conduct a population and family structure-corrected GWAS [@Chen2020; @Chen2016]. We visualized the results by constructing a Manhattan plot using snpR [@hemstromSnpRUserFriendly2022]. We identified the genes underlying any candidate regions by hand using the MonarchBase genome browser online tool [@Zhan2013].

@@

mod_AsRa <- lme(Aspect_Ratio ~ Yolked_Oocytes, random = ~ 1|Female_Line, old_phenotypic, na.action = na.omit)
mod_Area <- lme(Marea ~ Yolked_Oocytes, random = ~ 1|Female_Line, old_phenotypic, na.action = na.omit)

# Results
## Sequencing results
We sequenced a total of 417,846,352 reads across all individuals, 98.8% of which mapped to the monarch reference genome. 244,443,414 reads were maintained after filtering (1,158,500 per sample, on average), encompassing `r prettyNum(L,big.mark=",",scientific=FALSE)` total sequenced bases. From this, we called `r prettyNum(nsnps(dat),big.mark=",",scientific=FALSE)` individual SNPs following filtering.

@@

@@

flextable::regulartable(Tab1, cwidth = 1.2) %>% flextable::set_caption(., "Counts of individuals per reproductive status in each year.")
```

```{r Figure_1_print,echo=FALSE,warning=FALSE,fig.cap="Figure 1: Approximate ranges ancestral and introduced range with introduction dates for monarch butterflies in the Pacific, with sampling location noted in orange.",fig.width=8, fig.height=8, fig.align='center', fig.cap.style="Image Caption"}
Figure1 <- imager::load.image("../results/map_annotated.jpg")
plot(image, axes = FALSE)

>>>>

---
title: "Novel genetic control of migratory diapause in Australian Monarch Butterflies"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  word_document:
    reference_docx: "paper-files/style_ref.docx"
  pdf_document:
    keep_tex: true
    fig_caption: true
    latex_engine: pdflatex
    fig_width: 7
    fig_height: 6
bibliography: "paper-files/citations.bib"
bst: "paper-files/pnas-new.bst"
header-includes:
  - \usepackage[left]{lineno}
  - \linenumbers
  - \usepackage{setspace}\doublespacing
  - \usepackage{sectsty}\sectionfont{\fontsize{12}{12}\selectfont}
  - \usepackage{sectsty}\subsectionfont{\normalfont\itshape\fontsize{12}{12}\selectfont}
  - \usepackage[round]{natbib}
indent: true
sansfont: Times New Roman
fontsize: 12pt
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(snpR); library(data.table); library(ggplot2); library(gridExtra); library(dplyr); library(openxlsx); library(jpeg); library(nlme)

@@

# Results

## Background
Although the ancestral range of monarch butterflies is believed to be in North and Central America, they have fairly recently expanded into South America and the Caribbean, and very recently to several locations throughout the Atlantic and Pacific [@oplerFieldGuideWestern1999; @Zhan2014]. In North America, the species is well known for its unique, multi-generational migratory life history wherein individuals dispersing northward in the spring do so over three to four generations and then return to their wintering grounds in a single step [@malcolmSpringRecolonizationEastern1993; @cockrellTimeTemperatureLatitudinal1993].

@@

<!-- chunks located here to ensure everything calls correctly -->
```{r, read_sample_meta_get_numbers, echo=FALSE,message=FALSE,include=FALSE,warning=FALSE}
sample.meta <- read.table("../data/sample_metadata.txt", header = T, stringsAsFactors = F)
old.data <- data.table::fread("../data/Aus_2018_metadata.txt", header = T)
old.data[,fam := as.numeric(gsub("_.+", "", ID))]

@@

@@

# plot
Figure3b <- plot_manhattan(dat, "gmmat_pval_likely_pheno", chr = "scaffold", significant = 0.00005, suggestive = 0.0005, log.p = T,
                           chr_order = c(paste0("chr", 1:30), "other"), highlight_style = "red") # highlight top snp
Figure3b <- Figure3b$plot + ylab("-log10(p-value)") + xlab("Position") + theme(strip.text = element_blank())
```

@@

mod_AsRa <- lme(Aspect_Ratio ~ Yolked_Oocytes, random = ~ 1|Female_Line, old_phenotypic, na.action = na.omit)
mod_Area <- lme(Marea ~ Yolked_Oocytes, random = ~ 1|Female_Line, old_phenotypic, na.action = na.omit)
```

## Sequencing results
We incubated female monarch butterflies from a total of `r length(unique(na.omit(old.data$fam)))` clutches of eggs produced by females collected from a continously-breeding population from southern Queensland, Australia under a decreasing photoperiod following the methods of Freedman et al. [-@Freedman2018]. The resulting `r nrow(old.data)` females were phenotyped for oocyte development as a proxy for diapause status. `r sum(sample.meta$likely_pheno[sample.meta$year == 2018] == "R", na.rm = TRUE)` butterflies were phenotyped as reproductively mature (e.g. not in diapause, and thus a non-migratory phenotype) and `r sum(sample.meta$likely_pheno[sample.meta$year == 2018] == "B", na.rm = TRUE)`as reproductively immature (e.g. in diapause, and thus a migratory phenotype). We sequenced these individuals along with a additional `r sum(sample.meta$year == 2016, na.rm = TRUE)` butterflies raised under very similar conditions by Freedman et al. [-@Freedman2018], of which `r sum(sample.meta[sample.meta$year == 2016,]$likely_pheno == "R", na.rm = TRUE)` and `r sum(sample.meta[sample.meta$year == 2016,]$likely_pheno == "M", na.rm = TRUE)` were scored in that study as reproductively mature and immature, respectively.

We sequenced a total of 417,846,352 reads across all individuals, 98.8% of which mapped to the monarch reference genome. 244,443,414 reads were maintained after filtering (1,158,500 per sample, on average), encompassing `r prettyNum(L,big.mark=",",scientific=FALSE)` total sequenced bases. From this, we called `r prettyNum(nsnps(dat),big.mark=",",scientific=FALSE)` individual SNPs following filtering.

@@

# Methods
## Sample Collection, Incubation, and Phenotyping

In order to determine if any genomic regions were significantly associated with reproductive diapause, we used the snpR R package to filter out loci significantly out of Hardy-Weinburg Equilibrium ($p \le 0.000001$) in either year in order to help remove any paralogous loci and interpolated missing data by conducting binomial draws at missing genotypes using the minor allele frequencies at those loci [@hemstromSnpRUserFriendly2023]. To assess the basic level of genetic diversity amongst our samples, we then calculated the average number of pairwise differences ($\pi$), observed heterozygosity ($H_{O}$), and $F_{IS}$ according to Weir and Cockerham [-@Weir1984] for all of our samples and for each year independently using snpR [@hemstromSnpRUserFriendly2023]. We also calculated $F_{ST}$ according to Weir and Cockerham [-@Weir1984] between sample years and between both sample year and reproductive phenotype jointly [@hemstromSnpRUserFriendly2023]. We bootstrapped a significance level for $F_{ST}$ by bootstrapping individuals among populations with replacement 100 times. To determine if there was any substantial unexpected genetic structuring (other than family groups) in our samples, we conducted a Principal Component Analysis using snpR [@hemstromSnpRUserFriendly2023]. Since some individuals escaped their rearing dishes (but not the incubator) during rearing, we also confirmed the parentage of our samples using the Colony2 program, assuming polygamous, random mating [@Jones2010].

In order to determine the degree to which diapause status was biased between families, we constructed a Fisher's exact test for independence using 2000 simulations in R version 4.2.2 [@rcoreteamLanguageEnvironmentStatistical2022]. We then conducted a Genome-Wide Association Study (GWAS) in order to determine if any loci were significantly associated with reproductive status. Since our samples were composed of many groups of full and half-siblings and came from two genetically different years, we used the GMMAT R package to conduct a population and family structure-corrected GWAS [@Chen2020; @Chen2016]. We visualized the results by constructing a Manhattan plot using snpR [@hemstromSnpRUserFriendly2023]. We identified the genes underlying any candidate regions by hand using the MonarchBase genome browser online tool [@Zhan2013].

@@

flextable::regulartable(Tab1, cwidth = 1.2) %>% flextable::set_caption(., "Counts of individual monarch butterfly reproductive status in each year.")
```

```{r Figure_1_print,echo=FALSE,warning=FALSE,fig.cap="Figure 1: Approximate ancestral and introduced ranges with introduction dates for monarch butterflies in the Pacific, with sampling location noted in orange.",fig.width=8, fig.height=8, fig.align='center', fig.cap.style="Image Caption"}

# from https://stackoverflow.com/questions/9543343/plot-a-jpg-image-using-base-graphics-in-r
plot_jpeg <- function(path, add=FALSE)
{
  require('jpeg')
  jpg = readJPEG(path, native=T) # read the file
  res = dim(jpg)[2:1] # get the resolution, [x, y]
  if (!add) # initialize an empty plot area if add==FALSE
    plot(1,1,xlim=c(1,res[1]),ylim=c(1,res[2]),asp=1,type='n',xaxs='i',yaxs='i',xaxt='n',yaxt='n',xlab='',ylab='',bty='n')
  rasterImage(jpg,1,1,res[1],res[2])
}

plot_jpeg('../results/map_annotated.jpg') # needs to be rescaled a bit manually, or just insert the jpg directly into the docx

# Figure1 <- imager::load.image("../results/map_annotated.jpg")
# plot(image, axes = FALSE)
